Supplementary material

### 2022-01-14 analysis/paper/paper.Rmd
Trying to identify recommended offset for shoreline dating

```{r libraries, eval = TRUE, results = 'hide'}
library(here)
library(tidyverse)
library(sf)
library(ggridges)
library(ggthemes)
library(ggnewscale)
library(patchwork)
library(terra)
library(raster)
library(oxcAAR)
library(vapour)
library(topoDistance)
library(IRanges)
quickSetupOxcal()
```

@@

The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g. @berg-hansen2009; @bjerck2005; @sognnes2003; @solheim2018; @tallavaara2020].

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger in 1905 (see the review by @maccurdy1906 for an early mention in English). Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g. @berg-hansen2009; @bjerck1990; @mjærum2018]. The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

The archaeological data compiled for the analysis consists of all excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)). These number a total `r nrow(sites)` sites. Of these, `r length(unique(rcarbon$site_name))` sites are associated with the total of `r nrow(rcarbon)` radiocarbon dates. In turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE) with 95 % probability. These sites and ^14^C-dates form the basis for the analysis. Spatial data in the form of site limits and features, as defined by the excavating archaeologists, were retrieved from local databases at the Museum of Cultural History---the institution responsible for archaeological excavations in the region. In the compiled dataset, each radiocarbon date has been associated with the site features or excavation unit from where they originate, or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what available spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as that defined during initial survey, area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hull polygons generated around the site features.

Five of the sites have been associated with farming activities, either in directly or in the form building structures. The first is Nordby 1, which has . The Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are related to farming. Both sites also have ^14^C-dates and lithic inventory associated with Mesolithic forager activities. Finally, @nielsen2021 has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratberg and Lang could be related to early agricultural activity in the Oslo fjord region.

@@

# Results

@@

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. Following the general and aggregative result of this, one recommendation could be that given no other information, the method of shoreline dating should be applied with an uncertainty of 0-15m offset to the Stone Age site elevation relative to the displacement curves, following the median vertical distance and associated interquartile range. This would naturally be best applied when combined with a more comprehensive consideration of the site location, which could lead one to reduce this uncertainty, given for example surroundings where it would be reasonable to assume that a certain sea-levels would be unlikely in light of the surroundings of the site.

```{r results, echo = FALSE, fig.cap = "The result of running the analysis across all sites. Each data point is plotted with some transparency, meaning that the darker the colour, the more often those values occured. The first row shows the result of including all dates the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites. The second row shows the result of excluding these. The table under each plot lists some corresponding statistics for central tendency and dispersion.", out.width = "500px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/results.png"))

@@

```{r results2, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using only dates older than 2500 BCE that correspond to the site inventory.", out.width = "525px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/results2.png"))
```

@@

Some sources of likely variation and uncertainty have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, has not been taken into account. Secondly, the DTM has a vertical error... Thirdly, the tidal range has not been considered...  Fourthly, the exposure of the sites to wave action is also likely to have been of concern, and will presumably have implications for how close to the shoreline people settled. And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site. These dimensions will have implications for the shore-line dating for certain subsets of the sites and could be future avenues through which one could further improve the method. These dimensions will also be imprtant for more substantial considerations of the behavioural relation between sites and the shoreline.

>>>>

```{r libraries, eval = TRUE, results = 'hide'}
library(here)
library(tidyverse)
library(sf)
library(ggridges)
library(ggthemes)
library(ggnewscale)
library(patchwork)
library(terra)
library(raster)
library(oxcAAR)
library(vapour)
library(topoDistance)
library(IRanges)
library(stringr)
quickSetupOxcal()
```

@@

The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g. @berg-hansen2009; @bjerck2005; @sognnes2003; @tallavaara2020].

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger in 1905 (see the review by @maccurdy1906 for an early mention in English). Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g. @berg-hansen2009; @bjerck1990; @mjærum2018; @solheim2018]. The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

The archaeological data compiled for the analysis consists of all excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)). These number a total `r nrow(sites)` sites. Of these, `r length(unique(rcarbon$site_name))` sites are associated with a total of `r nrow(rcarbon)` radiocarbon dates. Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE) with 95 % probability. These sites and ^14^C-dates form the basis for the analysis. Spatial data in the form of site limits and features, as defined by the excavating archaeologists, were retrieved from local databases at the Museum of Cultural History---the institution responsible for archaeological excavations in the region. In the compiled dataset, each radiocarbon date has been associated with the site features or excavation unit from where they originate, or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what available spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as that defined during initial survey, area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hull polygons generated around the site features.

Three of the sites have been associated with farming activities, either in directly or in the form building structures. The first is Nordby 1, which has . Furthermore, the Middle Neolithic phase at Kvastad A2 and Late Neolithic phase at Nauen A are both related to farming. Both of these sites also have ^14^C-dates and lithic inventory associated with Mesolithic forager activities. These agricultural phases are thus treated seperately below.  Finally, @nielsen2021 has recently suggested that Early and Middle Neolithic features from the otherwise younger sites Bratberg and Lang could be related to early agricultural activity in the Oslo fjord region.

@@

```{r distances}
distances2 <- distances %>%
  filter(year <= -2500 & rcarb_cor == "t" & type == "forager") %>% 
  mutate(hordist = ifelse(str_detect(sitename, "Løvås"), 0, hordist),
         topodist = ifelse(str_detect(sitename, "Løvås"), 0, topodist),
         vertdist = ifelse(str_detect(sitename, "Løvås"), 0, vertdist))

distances3 <- distances %>%
  filter(year <= -4000 & rcarb_cor == "t" & type == "forager") %>% 
  mutate(hordist = ifelse(str_detect(sitename, "Løvås"), 0, hordist),
         topodist = ifelse(str_detect(sitename, "Løvås"), 0, topodist),
         vertdist = ifelse(str_detect(sitename, "Løvås"), 0, vertdist))
```
# Results

@@

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. The best point estimate for the distance between site and shoreline for the pre-Late Neolithic is, following from the skewed distribution, given by the median at `r round(median(distances2$vertdist))`m. However, 95% of the distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m. That is, for 95% of the cases, the shoreline was situated on or down to m below the site location. This is constrained further when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m. When applying the method of shoreline dating to Stone Age sites in the region, it could thus be recommended, given no other information, to allow for a potential deviation between site and shoreline from `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))` in vertical distance. This can be reduced to `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m for Mesolithic sites. The sites treated here have been given a shoreline date using these values in Figure. However,

Following the general and aggregative result of this, one recommendation could be that given no other information, the method of shoreline dating should be applied with an uncertainty of 0-15m offset to the Stone Age site elevation relative to the displacement curves, following the median vertical distance and associated interquartile range. This would naturally be best applied when combined with a more comprehensive consideration of the site location, which could lead one to reduce this uncertainty, given for example surroundings where it would be reasonable to assume that a certain sea-levels would be unlikely in light of the surroundings of the site.

Figure shows the result of shoreline dating the sites using the recommended off-set with radiocarbon dates plotted against this. The steeper the curve the more narrow the date range resulting from the procedure. This means that older siter, and sites situated towards the north-east will have a more narrow date range.

```{r results, echo = FALSE, fig.cap = "The result of running the analysis across all sites. Each data point is plotted with some transparency, meaning that the more intense the colour, the more often those values occured. Results associated with agricultural acitivites are in grey. The first row shows the result of including all dates the Stone Age, including those seen as otherwise  unrelated to the main occupation of the sites. The second row shows the result of excluding these. The table under each plot lists some corresponding statistics for central tendency and dispersion.", out.width = "500px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/results.png"))

@@

```{r results2, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using only dates older than 2500 BCE that correspond to the site inventory. Negative values have been set to zero.", out.width = "525px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/results2.png"))
```

```{r results3, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using only dates older than 4000 BCE that correspond to the site inventory. Negative values have been set to zero.", out.width = "525px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/results3.png"))
```

```{r shoredate, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites allowing for a sea-level potentially 4m above to 16m below the lowest site elevation. These are plotted against the oldest radiocarbon-informed dates for each site.", out.width = "525px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

@@

While allowing for a deviation up to 16m might be recommended when no other information is available, this can likely be considerably reduced by evaluating the surroundings of the site.

Some sources of likely variation and uncertainty have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, has not been taken into account. Secondly, the DTM has a vertical error... Thirdly, the tidal range has not been considered...  Fourthly, the exposure of the sites to wave action is also likely to have been of concern [@roalkvam2020], and will presumably have implications for how close to the shoreline people settled [@blankholm2020]. And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site. These dimensions will have implications for the shore-line dating for certain subsets of the sites and could be future avenues through which one could further improve the method. These dimensions will also be imprtant for more substantial considerations of the behavioural relation between sites and the shoreline.

### 2022-01-14 analysis/supplementary-materials/supplementary.Rmd
Trying to identify recommended offset for shoreline dating

This document shows the simulation results for each individual site analysed in the study with some additional notes pertaining to edits to the DTM, treatment of ^14^C-dates and other relevant information. The sites are presented in alphabetical order. Tables holding data on excavated sites and radiocarbon dates, and spatial data for site limits and site features are available in the online repoistory for the paper at https://osf.io/7F9SU

Sites with ^14^C-dates to the Stone Age, but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975], Nedre Holtan, Rugtvedt [@odgaard1994], Auve [@østmo2008] and Torsrød [@østmo1975]. These sites are therefore not included in the analysis.

@@

Prestemoen 1 has a lithic inventory consistent with the ^14^C-dates [@persson2014]. The site is situated by a gravel pit that is clearly visible on the DTM to the north-east of the site. This was not edited as it has not impacted the simulated distance between site and sea. The landscape surrounding the site would, however, have been substantially different in the Mesolithic [@persson2014, fig. 10.20].

@@

@@

Lithic inventory match the ^14^C-date from Rødbøl 54 [@mansrud2008].

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

Bratsberg was originally two sites [@wenn2012]. [@nielsen2021] has interpreted this as an indicator of early farming.

>>>>

This document shows the simulation results for each individual site analysed in the study with some additional notes pertaining to edits to the DTM, treatment of ^14^C-dates and other relevant information. The sites are presented in alphabetical order.  excavated sites and radiocarbon dates, and spatial data for site limits and site features are available in the online repoistory for the paper at https://osf.io/7F9SU

Sites with ^14^C-dates to the Stone Age but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975, although these do not typologically match], Nedre Holtan, Rugtvedt [@odgaard1994], Auve [@østmo2008] and Torsrød [@østmo1975]. These sites are therefore not included in the analysis. Furthermore, @nielsen2021 has recently interpreted features from the otherwise younger sites Bratsberg and Larønningen as possibly related to farming activity in the Early Neolithic. However, given the limited number of features, the lack of any artefacts directly related to these, and the therefore somewhat speculative nature of this suggestion, the sites are kept out of the analysis here.

@@

Prestemoen 1 has a lithic inventory consistent with the ^14^C-dates [@persson2014]. The site is situated by a gravel pit that is clearly visible on the DTM to the north-east of the site. This was not edited as it has not impacted the simulated distance between site and sea. The landscape surrounding the site would, however, have been very different in the Mesolithic [@persson2014, fig. 10.20].

@@

The earliest ^14^C-date does not match the Middle Mesolithic inventory from Ragnhildrød [@mjaerum2012]. The post-hole dated to the Late Neolithic/Bronze Age could be related to a single find of a pressure flaked object, but was not given much weight in the report [@mjaerum2012, 76].

@@

Lithic inventory match the ^14^C-date from Rødbøl 54 [@mansrud2008]. The site is situated where the highway runs today, which required editing.

@@

Rognlien was originally excavated in [@ingstad1970]. The site limit has been manually defined based on the description given in the report and the spatial geometry of the site as defined in the database Askeladden. It could be slightly off, but should match the elevation given in the publication of the site [@ingstad1970].The ^14^C-dates from the original excavation was not seen as contemporaneous with the lithic material on the site. Burnt elk bone from the original excavation was therefore recently subjected to radiocarbon dating and is believed to match the occupation of the site [pers.comm. Steinar Solheim, also provided in @stokke2017, 24].

@@

^14^C-dates from Sagene B1 do not match the Early Mesolithic inventory of the site [@viken2018]. The wide discrepancy of the dates from the the post-holes also led @viken2018 to suggest that these features might actually be non-anthropogenic and rather be related to trees that have grown on the site.

@@

The ^14^C-dates from Sagene B2 do not match the clearly Early Mesolithic inventory of the site [@darmark2018a].

@@

Solum 3 is a Late Neolithic and Early Bronze Age site, with ^14^C-dates matching the lithic inventory [@fossum2014d].

@@

The artefact inventory from Stokke/Polland 1, containing both lithics and ceramics, indicates visits in the Late Mesolithic as well as the Early Neolithic [@koxvold2017i]. Date groups were therefore manually separated, even though they overlap at 99.7% probability. The site is located right by the highway to the east, which was edited for the other Stokke/Polland sites, but this does not appear relevant for the reconstruction of the sea-level at Stokke/Polland 1, which would have reached the site from the west.

@@

The ^14^C-dates matches the lithic inventory from Stokke/Polland 5 [@mansrud2017]. The site is located where the highway runs today. This appears to have been successfully edited. The excavated area is more extensive (field A--C in the report) that what is used as the site limit here, but the relation between the different areas is not clear. Field A, used as the site limit here, holds all the dated features.

@@

Stokke/Polland 8 have corresponding lithic inventory and ^14^C-date [@fossum2017c]. The site is situated just east of where the highway runs today. This appears to have been successfully edited [cf. @fossum2017c, fig.28.3].

@@

Typology of lithics match the ^14^C-dates from Torstvet [@mansrud2013b]. The site is located where the highway runs today. This appears to have been successfully edited, although there is a possibility that this has led to the sea-level being simulated sightly further away from the site than what would otherwise have been the case. However, as with he Hovland-sites above, this appears to rather follow from the general topography of the landscape which makes the reconstruction more uncertain than in areas with a higher relief.

@@

^14^C-dates match the typological indicators in the artefact assemblage at Trondalen [@mansrud2018c]. No disturbances required editing in the DTM.

@@
